---
title: "seqFISH Spatial Data Workflow"
author: "Dario Righelli"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{seqFISH Spatial Data Workflow}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading data and libraries

We stored the provided seqFish and scRNA-seq data inside the SpatialExperiment
external data seqFISH folder.

```{r}
library(SpatialExperiment)
```


## seqFISH data

```{r}
fishCoordFile <- system.file(file.path("extdata", "seqFISH",
                            "fcortex.coordinates.txt"), 
                            package="SpatialExperiment")
fishCoordinates <- read.table(fishCoordFile, header=FALSE, sep=" ")
colnames(fishCoordinates) <- c("ID", "Irrelevant", "x", "y")

fishCellLabsFile <- system.file(file.path("extdata", "seqFISH", 
                            "seqfish_cell_labels.tsv"),
                            package="SpatialExperiment")
fishCellLabels <- read.table(file=fishCellLabsFile, header=FALSE, sep="\t")
colnames(fishCellLabels) <- c("ID", "cluster", "class", "classID", 
                                "Irrelevant", "Prob")
fishFeatCountsFile <- system.file(file.path("extdata", "seqFISH",
                            "seqfish_normalized_cortex_b2_testing.txt"), 
                            package="SpatialExperiment")
fishFeaturesCounts <- read.table(file=fishFeatCountsFile, 
                                header=FALSE, sep="\t", row.names=1)
```

## Single Cell RNA-seq data

```{r}
scRNACountsFile <- system.file(file.path("extdata", "seqFISH", 
                            "tasic_scRNAseq_training_b2.txt"),
                            package="SpatialExperiment")
scRNASeqCounts <- read.table(file=scRNACountsFile, header=FALSE, 
                            sep="\t", row.names=1)
scRNASeqCounts <-  as.matrix(scRNASeqCounts)
scRNALabsFile <- system.file(file.path("extdata", "seqFISH", 
                            "tasic_scRNAseq_labels.tsv"),
                            package="SpatialExperiment")
scRNASeqLabels <- read.table(file=scRNALabsFile, header=FALSE, sep="\t")
```

# SingleCellExperiment package

We load the scRNA-seq data into a SingleCellExperiment class, so we can use it
for further joint analysis with the seqFISH data.

```{r}
sce <- SingleCellExperiment(rowData=rownames(scRNASeqCounts), 
                    colData=scRNASeqLabels,
                    assays=SimpleList(counts=as.matrix(scRNASeqCounts)))
show(sce)
```


# SpatialExperiment package

<!-- ## Installation -->

<!-- We install the SpatialExperiment package which is available through the github/drighelli -->
<!-- repository on spatialExperiment branch. -->

<!-- ```{r} -->
<!-- library(devtools) -->

<!-- install_github("drighelli/SpatialExperiment", ref="spatialExperiment") -->
<!-- ``` -->

## Class handling

We can load the data inside the SpatialExperiment class.
We then use the show method for looking at the class description.

```{r}
se <- SpatialExperiment(rowData=rownames(fishFeaturesCounts),
                        colData=fishCellLabels,
                        assays=SimpleList(counts=as.matrix(fishFeaturesCounts)),
                        spatialCoords=fishCoordinates)
show(se)
```

## Spatial Coordinates

With the aid of the spatialCoords methods we can get and set the spatial coordinates.

### Getter

```{r}
spatialCoords(se)
```

### Setter

We create a fake fish coordinates data frame and overwrite the old one,
showing that the y coordinates are identical to the x ones.

```{r}
fakeFishCoords <- cbind(fishCoordinates[,c(1:3)], fishCoordinates[,3])
colnames(fakeFishCoords) <- c("ID", "Unknown", "x", "y")
spatialCoords(se) <- fakeFishCoords
spatialCoords(se)
```

