---
title: "10x-Visum Spatial Data Analysis"
author: "Dario Righelli"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    BiocStyle::html_document:
      toc: true
vignette: >
    %\VignetteIndexEntry{10x-Visum Spatial Data Analysis}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installation

Installing and loading VisualExperiment class.

<!-- ```{r} -->
<!-- library(devtools) -->
<!-- install_github("drighelli/SpatialExperiment", ref="spatialExperiment") -->
<!-- ``` -->


```{r}
library("SpatialExperiment")
library("Matrix")
library("rjson")
```


# Data Loading

Examples data files have been downloaded from the 10x site, [Mouse adult brain](https://support.10xgenomics.com/spatial-gene-expression/datasets/1.0.0/V1_Adult_Mouse_Brain)

## Features

Loading features and count matrix.

```{r}
barcodesFile <- system.file(file.path("extdata", "10x_visium",
                            "barcodes.tsv"), package="SpatialExperiment")
barcodesEx <- read.csv(barcodesFile, sep="\t", 
                     header=FALSE, col.names=c("Barcodes"))
featuresFile <- system.file(file.path("extdata", "10x_visium",
                            "features.tsv"), package="SpatialExperiment")
featuresEx <- read.csv(featuresFile, sep="\t", 
                     header=FALSE, col.names=c("Barcodes", "Feature_name", 
                                               "Feature_type"))

countsFile <- system.file(file.path("extdata", "10x_visium",
                            "matrix.mtx"), package="SpatialExperiment")
countsEx <- readMM(file=countsFile)

```

## Spatial Coordinates

Loading Spatial coordinates of the tissue section.

```{r}
posFile <- system.file(file.path("extdata", "10x_visium",
                        "tissue_positions_list.tsv"), 
                        package="SpatialExperiment")
tissPosEx <- read.csv(posFile, 
                        sep="\t", header=FALSE, 
                        col.names=c("Barcodes", "in_tissue", 
                                    "array_row", "array_col",
                                    "pxl_col_in_fullres", "pxl_row_in_fullres"))
```

## Spatial Scale Factors

Scale factors for the given three image formats transformations.

```{r}
scaleFile <- system.file(file.path("extdata", "10x_visium",
                        "scalefactors_json.json"), 
                        package="SpatialExperiment")

scalefactors <- fromJSON(file=scaleFile)
```


# VisiumExperiment

## Creation

Creating VisiumExperiment class within 10x Visium data and testing the show method.

```{r}
ve <- VisiumExperiment(rowData=featuresEx, colData=barcodesEx, 
                           assays=c(counts=countsEx), 
                           spatialCoords=tissPosEx,
                           scaleFactors=scalefactors)
ve
```

## Methods

### Retrieving Spatial Coordinates

The *spatialCoords* methods can be used to retrieve the spatial coordinates., 
It can be used also to update the values.

```{r}
spatialCoords(ve)
```

### Number of array barcoded-spots covered by tissue

The *isInTissue* method can be used to know which and how many spots are covered
by the tissue.

```{r}
sum(isInTissue(ve))
# ve[isInTissue(ve)] ## subset not defined yet
```

```{r}
sessionInfo()
```

