---
title: "Demo"
author: "Helena L. Crowell"
date: "`r format(Sys.Date(), '%b %m, %Y')`"
output: BiocStyle::html_document
---

<style type="text/css"> .smaller { font-size: 10px } </style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, cache.lazy = FALSE)
```

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(scater)
library(SpatialExperiment)
library(tidyr)
```

# Summary

A brief summary of the package's current structure is outlined below. To give some examples, we load an example `SpatialExperiment` containing 10X Visium spatial gene expression data of two serial mouse brain sections (Sagittal-Posterior) available [here](https://support.10xgenomics.com/spatial-gene-expression/datasets).

```{r}
data(ve)
```

## The `SpatialExperiment` class {- .tabset}

### Spatial data {-}

Spatial data are stored in the `colData` and must contain:

* a character `sample_id` column giving unique sample identifiers
* a logical `in_tissue` column indicating whether or not an observation was mapped to tissue
* a two- or three-column numeric `xy_coords` matrix storing spatial coordinates
  
```{r}
head(colData(ve))
```
  
### Image data {-} 
  
Image-related data are stored in the `int_metadata()$imgData` field as a `DataFrame` with the following columns:

* `sample_id` and `image_id` specifying the image's sample and image identifier
* `data`: a list of `SpatialImage`s containing the image's `grob`, path and/or URL
* `width` and `height` giving the image's dimension in pixel
* `scaleFactor` used to re-scale spatial coordinates according to the image's resolution 

```{r}
(df <- imgData(ve))
```

### Getters, setters & methods {-}

* `scaleFactors()` retrieves scale factors
* `imgData(), imgData() <- value` to get/set the `imgData` `DataFrame`
* `add/removeImg()` can be used to create/eliminate images in the `imgData` 
* `load/unloadImg()` can be used to load/unload images, i.e. add/remove the `grob` 
* the `read10xVisium()` constructor can be used to create a `SpatialExperiment` from 10x Visium spatial gene expression data in one line

## The `SpatialImage` class

Contains three slots that store any available information associated with an image: 

* `@grob`: NULL or an object class `rastergrob` from the `grid` package 
* `@path`: NULL or a character strings specifying an image file name (.png, .jpg or .tif)
* `@url`: NULL or a character string specifying an URL from which to retrieve the image

```{r}
df$data[[1]]
```

* `imgGrob/path/Url(), ... <- value` access/set data in the respective slots
* `load/unloadImg()` are used to add/drop the `grob` slot
* `loadImg()` supports caching when loading from a URL; in general, paths are given precedence over URLs

# The `SpatialExperiment` class

The `SpatialExperiment` class extends the `SingleCellExperiment` class by requiring specific fields to be present in the object's `colData` and `int_metadata`. These aim to accommodate spatially and image related data", "respectively. 

# Reading 10X Visium data

The 10X Genomics' *CellRanger* pipeline will process the data using standard output file formats that are saved, for each sample, in a single directory `/<sample>/outs/` of the following structure: 

```{bash eval = FALSE}
sample 
|—outs 
··|—raw/filtered_feature_bc_matrix.h5 
··|—raw/filtered_feature_bc_matrix 
····|—barcodes.tsv 
····|—features.tsv 
····|—matrix.mtx 
··|—spatial 
····|—scalefactors_json.json 
····|—tissue_lowres_image.png 
····|—tissue_positions_list.csv 
```

We can load these data into a `SpatialExperiment` using the `read10xVisium` function", "which will read in all relevant information", "including the count data", "spatial coordinates", "scale factors", "and images:

```{r}
dir <- system.file(
  file.path("extdata", "10xVisium"), 
  package = "SpatialExperiment")

sample_ids <- c("section1", "section2")
samples <- file.path(dir, sample_ids)

list.files(samples[1])
list.files(file.path(samples[1], "spatial"))
list.files(file.path(samples[1], "raw_feature_bc_matrix"))

(ve <- read10xVisium(samples, sample_ids,
  images = "lowres", # specify which image(s) to include
  load = TRUE))      # specify whether or not to load image(s)
```

# Spatial data

Sample identifiers, indication of whether or not an observation was mapped to the tissue, as well as spatial coordinates are stored inside the `colData`.

```{r}
# tabulate number of spots mapped to tissue
table(
  in_tissue = ve$in_tissue,
  sample_id = ve$sample_id)
```

```{r}
# view spatial coordinates
head(ve$xy_coords)
```

# Image data

All image related data are stored inside the `int_metadata`'s `imgData` field as `DataFrame` of the following structure: 

* each row corresponds to one image for a given sample and with a given unique image identifier (e.g. its resolutions)
* for each image, columns specify:
  * which `sample_id` the image belongs to
  * a unique `image_id` in order to accommodate multiple images for a given sample (e.g. of different resolutions)
  * the image's `data` (a `SpatialImage` object)
  * `width` and `height` of the image in pixel 
  * the `scaleFactor` that adjusts pixel positions of the original, full-resolution image to pixel positions in the image

The `imgData()` accessor can be used to retrieve the image data stored within the object:

```{r}
imgData(ve)
```

## The `SpatialImage` class

Images are stored inside the `data` field of the `imgData` as a list of `SpatialImage`s, which enables storing three types of information that may be associated with an image:

* `grob`: a `grob` object of the image 
* `path`: a file path from which to load the image
* `url`: a URL from which to retrieve the image

Data available in an object of class `SpatialImage` may be accessed via the `imgGrob()`, `imgPath()` and `imgUrl()` accessors.

```{r}
imgData(ve)$data
```

While `grob`s can be used directly for plotting (e.g. using `grid.draw` or `ggplot2`), when multiple images are to be stored (say, for many samples and of different resolutions), or when a `SpatialExperiment` is to be exported, the `path` and `url` provide the option to store an image's source at minimal storage cost. 

```{r fig.small = TRUE}
# retrieve 'SpatialImage' for 1st image entry & visualize it
si <- imgData(ve)$data[[1]]
grid::grid.draw(imgGrob(si))
```

## Methods for image handling

The `SpatialExperiment` package provides various functions to handle which and how image data is stored in the object. These include:

* `loadImg` to actively load the image (from a path or URL) and store it as a `grob`
* `unloadImg` to drop the `grob`, while retaining the source path and/or URL
* `addImg` to add an image entry (as a path, URL, or `grob`)
* `removeImg` to drop an image entry entirely

Besides a path or URL to source the image from and a numeric scale factor, `addImg()` requires specification of the `sample_id` the new image belongs to, and an `image_id` that is not yet in use for that sample:

```{r fig.small=TRUE}
url <- "https://i.redd.it/3pw5uah7xo041.jpg"
ve <- addImg(ve, 
  sample_id = "section1", image_id = "pomeranian",
  imageSource = url, scaleFactor = NA_real_, load = TRUE)

grb <- imgGrob(ve, 
  sample_id = "section1", 
  image_id = "pomeranian")
grid::grid.draw(grb)
```

`loadImg()` and `add/removeImg()` are more flexible in the specification of the `sample/image_id` arguments. Specifically, 

- `TRUE` is equivalent to *all*, e.g. `sample_id = "<sample>", image_id = TRUE` will drop all images for a given sample.
- `NULL` defaults to the first entry available, e.g., `sample_id = "<sample>", image_id = NULL` will drop the first image for a given sample.

For example, `sample_id,image_id = TRUE,TRUE` will specify all images; `NULL,NULL` corresponds to the first image entry in the `imgData`; `TRUE,NULL` equals the first image for all samples; and `NULL,TRUE` matches all images for the first sample.

**Example 1:** Unload all images, i.e., drop all `grob`s. As a result, `grob` slots will be set to `NULL`, and all `SpatialImage`s now say `> not loaded`.

```{r}
imgData(ve <- unloadImg(ve, sample_id = TRUE, image_id = TRUE))$data
```

**Example 2:** Reload the first image for sample `section2`; the corresponding image now says `grob: Av` (for available) and `> loaded`.

```{r}
imgData(ve <- loadImg(ve, "section2"))$data[[2]]
```

**Example 3:** Remove `lowres` image for `section1` sample; the image is now completely gone from the `imgData`.

```{r}
imgData(ve <- removeImg(ve, "section1", "pomeranian"))
```

## Image caching

In general, paths take precedence over URLs, i.e. an image will only be loaded from the `imgUrl` if there is no `imgPath` available.

Images loaded from URLs will be cached, i.e. they are downloaded once and the file path where they were stored in will be stored under the `SpatialImage`'s `imgPath`:

```{r results="show"}
imgData(loadImg(ve, TRUE, TRUE))$data[[2]]
```

For example, if we unload `section1`'s `fullres` image now and reload it, there won't be another progress bar (the chunk somehow above doesn't display one anyways... but it does in the console):

```{r}
ve <- unloadImg(ve, TRUE, TRUE)
ve <- loadImg(ve, TRUE, TRUE)
```

# `colData` replacement

While storing of `sample_id`s, the `in_tissue` indicator, and spatial `xy_coords` inside the `SpatialExperiment`'s `colData` enables directly accessibility via the `colData` and `$` accessors, these fields are protected against arbitrary modification. This affects replacement operations to the following effects:

Renaming is generally not permitted:

```{r error = TRUE}
names(colData(ve))[1] <- "a"
```

Replacement of `sample_id`s is permitted provided that 

i. the number of unique sample identifiers is retained
ii. newly provided sample identifiers are a one-to-one mapping

```{r error = TRUE}
ve$sample_id <- sample(c("a", "b", "c"), ncol(ve), TRUE)
ve$sample_id <- sample(c("a", "b"), ncol(ve), TRUE)
```

Valid replacement will be propagated to the `imgData`:
  
```{r}
tmp <- ve
i <- as.numeric(factor(ve$sample_id))
tmp$sample_id <- c("sample1", "sample2")[i]
imgData(tmp)
```

The `in_tissue` and `xy_coords` fields may be modified provided that the former is a logical vector, and the latter is a two- or three-column numeric matrix:

```{r error = TRUE}
ve$in_tissue <- "x"
ve$xy_coords <- "x"
```

`colData(.) <- NULL` will retain only the `sample_id`, `in_tissue` and `xy_coords` fields:

```{r}
names(colData(ve))
colData(ve) <- NULL
names(colData(ve))
```

# Visualization

The `DataFrame` format enforced by the `imgData` field works nicely with `ggplot2`. Here's a prototype of what a wrapper for plotting could look like that supports

- coloring by observation metadata or features
- highlighting according to a logical statement
- faceting by multiple samples and/or features
- controlling whether to show points and/or images
- controlling which samples and images to use

```{r}
ggspatial <- function(x, 
  fill=NULL, highlight=NULL, assay="logcounts", 
  sample_id=NULL, image_id=NULL, points=TRUE, images=TRUE) 
{
  # check validity of input arguments
  stopifnot(
    is.logical(points), length(points) == 1,
    is.logical(images), length(images) == 1)
    
  if (is.null(assay)) {
    assay <- assayNames(x)[1]
  } else {
    stopifnot(
      is.character(assay), 
      length(assay) == 1, 
      assay %in% assayNames(x))
  }
  
  if (is.null(sample_id)) {
    # default to using all samples
    samples <- unique(x$sample_id)
  } else {
    stopifnot(sample_id %in% x$sample_id)
    samples <- sample_id
  }
  # subset specified samples
  x <- x[, x$sample_id %in% samples]
  
  if (is.null(image_id)) {
    # default to first available image for each sample
    idx <- SpatialExperiment:::.get_img_idx(x, TRUE, NULL)
  } else {
    if (length(image_id) == 1) {
      idx <- SpatialExperiment:::.get_img_idx(x, TRUE, image_id)
    } else {
      stopifnot(length(image_id) == length(samples))
      idx <- mapply(s=samples, i=image_id,
        function(s, i) .get_img_idx(x, s, i))
    }
  }
  # subset specified images
  imgData(x) <- imgData(x)[idx, ]
  
  df <- as.data.frame(colData(x))
  xy <- grep("xy", names(df))
  xy <- names(df)[xy] <- c("x", "y")
  
  # scale spatial coordinates
  sfs <- setNames(scaleFactors(x, TRUE, TRUE), samples)
  df[, xy] <- df[, xy] * sfs[df$sample_id]
  
  facets <- "sample_id"
  if (!is.null(fill)) {
    stopifnot(
      is.character(fill), 
      all(fill %in% rownames(x)) 
      || length(fill) == 1 & !is.null(x[[fill]]))
    if (all(fill %in% rownames(x))) {
      y <- assay(x, assay)
      y <- as.matrix(y[fill, ])
      df <- cbind(df, t(y))
      df <- pivot_longer(df, all_of(fill))
      fill <- "value"
      facets <- c(facets, "name")
    }
  }
  
  # in principle, images can be plotted using 'annotation_custom()'
  # however, this does not allow for faceting and we instead 
  # construct a separate image layer for each sample
  if (images) {
    # split 'imgData' by sample
    dfs <- split(imgData(x), imgData(x)$sample_id)
    # construct separate image layer for each sample
    images <- lapply(dfs, function(.) layer(
      data=as_tibble(.),
      inherit.aes=FALSE,
      stat="identity",
      position="identity",
      geom=ggplot2::GeomCustomAnn,
      params=list(
        grob=imgGrob(.$data[[1]]),
        xmin=0, xmax=.$width,
        ymin=0, ymax=.$height)))
  } else images <- NULL
  
  if (points) {
    guide <- ifelse(is.numeric(df[[fill]]), guide_colorbar, guide_legend)
    points <- list(
      guides(fill=guide(
        title=ifelse(fill == "value", assay, fill), 
        order=1, override.aes=list(col=NA, size=3))),
      geom_point(shape=21, size=1.25, stroke=0.25))
    if (!is.null(highlight)) {
      txt <- sprintf("with(df,%s)", highlight)
      df$highlight <- eval(parse(text=txt))
      highlights <- list(
        scale_color_manual(highlight, values=c("transparent", "black")),
        guides(col=guide_legend(override.aes=list(
          size=2, stroke=1, col=c("grey", "black")))))
    } else {
      df$highlight <- "transparent"
      highlights <- scale_color_identity()
    }
  } else {
    # this is required, else the image layer doesn't show
    points <- geom_point(col="transparent")
    highlights <- NULL
  }

  if (!is.null(facets)) {
    if (length(facets) == 1) {
      facets <- facet_wrap(facets)
    } else {
      ns <- vapply(df[facets], function(.) length(unique(.)), numeric(1))
      if (ns[2] > ns[1]) facets <- rev(facets)
      facets <- reformulate(facets[1], facets[2])
      facets <- list(
        facet_grid(facets, switch="y"),
        theme(strip.text.y.right = element_text(angle=90)))
    }
  } else facets <- NULL
  
  ggplot(df, 
    aes_string("x", "y", fill=fill, col="highlight")) +
    images + points + highlights + facets +
    coord_equal(expand=FALSE) + theme_void() +
    theme(legend.key.size=unit(0.5, "lines"))
}
```

Here are some example visualizations with `ggspatial()`:

```{r}
# read10xCounts gives a 'DelayedMatrix'
# which doesn't work with 'scater'
assay(ve) <- as(assay(ve), "matrix")

# normalization & quality control
ve <- logNormCounts(ve)
ve <- addPerFeatureQC(ve)
ve <- addPerCellQC(ve)
```

```{r fig.width=6, fig.height=3}
# plotting images only
ggspatial(ve, points=FALSE)
```

```{r fig.width=7, fig.height=3}
# both samples colored by total counts with 'in_tissue' spots highlighted
ggspatial(ve, fill="total", highlight="in_tissue", images=FALSE) + 
  scale_fill_viridis_c(trans="log10")
```

```{r fig.width=12, fig.height=5.5}
# expression of some arbitrary highly expressed genes across samples
gs <- rownames(ve)[rowData(ve)$mean > 250]
ggspatial(ve[, ve$in_tissue], fill=gs) + scale_fill_viridis_c()
```

# Session information {- .smaller}

```{r session-info}
sessionInfo()
```

