---
title: "seqFISH Spatial Data Workflow"
author: "Dario Righelli"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{seqFISH Spatial Data Workflow}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    chunk_output_type: inline
---

<style type="text/css"> .smaller { font-size: 10px } </style>

# Introduction

The SpatialExperiment package provides classes and methods for single cell 
spatial data handling.

This vignettes shows how to create a SpatialExperiment class by loading a 
seqFISH spatial dataset.

The seqFISH data are part of the [BIRS Biointegration Workshop Hackathon](https://github.com/BIRSBiointegration/Hackathon/tree/master/seqFISH) 
previously published in [Zhu et al. 2018](https://www.nature.com/articles/nbt.4260).


# Installation 

```{r, eval=FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpatialExperiment")
```


# Loading data and libraries

We stored the provided seqFish and scRNA-seq data inside the SpatialExperiment
external data seqFISH folder.

```{r message = FALSE}
library(ggplot2)
library(SpatialExperiment)
```

# Reading seqFISH data

```{r}
dir <- system.file(
    file.path("extdata", "seqFISH"), 
    package = "SpatialExperiment")

counts <- read.table(
    file.path(dir, "seqfish_normalized_cortex_b2_testing.txt"),
    header = FALSE, row.names = 1, sep = "\t")

labels <- read.csv(
    file.path(dir, "seqfish_cell_labels.tsv"),
    header = FALSE, sep = "\t", col.names = c(
        "cell_id", "cluster", "class", 
        "class_id", "irrelevant", "prob")) 

coords <- read.table(
    file.path(dir, "fcortex.coordinates.txt"),
    header = FALSE, sep = " ", col.names = c(
        "cell_id", "irrelevant", "x", "y"))
```

```{r fig.width=6, fig.height=3}
# construct 'SpatialExperiment'
cd <-  DataFrame(
    sample_id = "cortex",
    in_tissue = TRUE,
    xy_coords = I(as.matrix(coords[, c("x", "y")])),
    labels[, c("cluster", "class", "class_id", "prob")])

(fse <- SpatialExperiment(
    assays = list(counts = counts), 
    colData = cd))

df <- as.data.frame(colData(fse))
xy <- grep("xy", names(df))
names(df)[xy] <- c("x", "y")
ggplot(df, aes(x, y, col = class)) + 
    geom_point() + coord_equal()
```

# Session information {- .smaller}

```{r session-info}
sessionInfo()
```

